Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-04-25 Manuscript/PLoSONEresponse/response.tex

\textcolor{blue}{The bias in the analysis in favor of Bayesian methods makes the recommendation to use Bayesian methods less satisfying.}

Our results did indeed biased us in favour of using a Bayesian approach to fix topology since our results clearly indicate that it is less likely to generate a wrong topology using a Bayesian approach (see Figure 4 and 5 and discussion line 472 to 496 in the first submission manuscript).
% TG: or I'm not sure I understood that one.
%-------------

\textcolor{blue}{The use of Mk rather than Mkv makes the study less relevant to real data analyses.}

The Mk model assumes that some morphological characters \textbf{can} be constant and the Mkv model assumes that \textbf{all} the characters are variable. Although the reviewer is right that real data analyses do not contain invariable (constant) morphological characters, we designed our simulations to be a generalisation of real data. Because we sampled the morphological characters rate from a gamma distribution with $\alpha$=0.5, some morphological characters had only a few state changes accross the matrix. When removing the data, there was a probability for these characters to be constant (e.g. if all the taxa in the matrix appart from one had a character state 0 for a given character and only one taxa had a state 1 for the same character, if that taxa had no data available for that character, then the character appeared to be constant in the matrix). Note that both RAxML and MrBayes do not distinguish between both models since the Mkv model is just a particular case of the Mk model and are not distinguished either in the original publication that we cite in the manuscript (Lewis 2001). We've changed the mentions of the M\textit{k} model to M\textit{kv} model throughout the new manuscript.

%Add note that we changed it along the text. RAxML and MrBayes are using default.
%-------------

@@

We changed ``infer" into ``simulate" both at line 88 and in the Figure 1 caption.

%-------------

\textcolor{blue}{line 112: does diversitree just stop simulating when it reaches the correct number of tips, or does it deal with the biases that can result from doing that? (Stadler 2011 is the ref for a discussion of this issues, I believe)}

See our response to the comment ``The procedure for generating a tree with fossils is a valid procedure." below.

%-------------

@@

We are not entirely sure if the reviewers refers to a sampling bias or to the problem link to constant characters described in Lewis (2001) (acquisition bias). If the reviewers are referring to the second (acquisition bias), both phylogenetic inference software we used (RAxML and MrBayes) are correcting for the this bias (-K MK  see Supplementary information 1 for the full list of commands). Also, the effect of the acquisition bias is to inflate branch length which is an aspect of the phylogeny we didn't explored in this study (focusing on topology). See response below to the comment \color{blue}{\textit{The use of Mk rather than Mkv makes the study less relevant to real data analyses.}}.

%-------------

\textcolor{blue}{line 171 - 177: the authors need to explain the order of operations for the $M_F$ vs $M_C$ masking. It makes a difference. If you mask out $M_F$ percentage of cells, and then mask $M_C$ percentage of columns, it is possible (indeed very probable) that the percentage of missing data in fossil characters that remain will not be $M_F$ after some of the morphological columns are removed. If you mask by $M_C$ first, and then $M_F$, the resulting matrices will always have $M_F$ percentage missing data (modulo rounding error) in the simulated realizations.}

We clarified this point ($M_C$ comes second) in the new version of the manuscript. Also, we clarified the potential confusion between the amount of data removed in one parameter ($M_F$ = 10\% for example) and the amount of missing data in the morphological matrix (5\%).

Lines 171 to 178 in the previously submitted manuscript:

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $M_C$ removes columns across both living and fossil taxa's data. Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.
\end{minipage}

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $M_C$ removes columns across both living and fossil taxa's data. Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa. We first applied the parameters $M_L$ and $M_F$ on the matrix and then applied the $M_C$ parameter. Therefore, when 10\% data was missing for both $M_L$ and $M_F$, 10\% data was missing in the morphological part of the matrix. However, when applying the $M_C$ parameter with the same rate (10\%) the resulting matrix potentially had more than 10\% data missing.
\end{minipage}

%-------------

@@

%-------------

\textcolor{blue}{Header for 3. should be ``building phylogenetic estimates" or ``estimating phylogenies" because the previous section described how the phylogenies and character data were ``built."}

We changed the Header 3 to ``estimating phylogenies". %change in manuscript

%-------------

@@

%Probably no impact since it's not using an MCMC. But should we try it? (approx. one to two (max) month calculation).

%-------------

\textcolor{blue}{Filtering the data sets for strong bootstrap support under the full data analysis surely introduces some bias in favor of the no-missing data analyses.}

We implemented this step of keeping trees with a minimum total median bootstrap to make the topological comparisons more robust. If the no-missing data tree has low node support, it is likely that the changes in topology for the missing data trees are not linked to our focal question (the effect of removing characters, living taxa or fossils) but just to the fact that low supported nodes can be just random. We've added the following justification in the new manuscript:

Lines 208 to 210 in the previously submitted manuscript:

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete" and 50 x 125 ``missing-data" matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50).
\end{minipage}

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete" and 50 x 125 ``missing-data" matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50). This step was implemented to make sure that the differences we observed in topologies (see below) where due to the amount of missing data for each parameters ($M_L$, $M_F$ and $M_C$) and not simply to low branch support that is likely to lead to different topologies.
\end{minipage}

%-------------

@@

%-------------

\textcolor{blue}{MrBayes: using strongly informative priors for the true parameters does more than speed up the searching, it represents an ``unfair" advantage (an option that would not be availble in real data analysis in which one does not know the true values) for the Bayesian method. This makes the studies report of superior behavior of Bayesian methods less persuasive.}

We would like to emphasize that in this study we focused on the effect of missing data in the morphological part of the matrix on topology only.The choice of the following priors (``exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions" (line 228) and the ``transition/transversion ratio prior of two sampled from a strong beta 229 distribution ($\beta$(80,40))" (line 229)) affected mainly the molecular part of the matrix that remains constant (gamma distribution and transition/transversion ratio) and where a minor part of the MCMC sampler (2.17\% of move probability for the three priors). Therefore we are confident that the topology is not affected allowing us to explore the effect of missing morphological data on topology as well as reducing our computational time to a realistic 1.5 CPU century.

%-------------

### 2015-04-25 Manuscript/PLoSONEresponse/TEM_PLoS_response.tex
Updated response

\section*{Methods}
To explore how missing data in the morphological sections of Total Evidence matrices influences tree topology, we used the following protocol (note that we explain each step in detail below this general outline; Fig.~\ref{Fig_Outline}).
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true" tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete" matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). We call the resulting 125 matrices ``missing-data" matrices.
\item{Inferring phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete" matrix and from the 125 ``missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the ``best" tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data" trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best" tree to the ``missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

@@

In practice, each parameter represents a different way of removing data from the matrix: $M_{L}$ removes rows from the living taxa's data; $M_{F}$ removes cells from the fossil taxa's data; and $M_{C}$ removes columns across both living and fossil taxa's data. Note that $M_{L}$ and $M_{F}$ differ not only because of the region of the matrix affected: for $M_{L}$ all the morphological data of a percentage of living taxa are removed, whereas for $M_{F}$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.

@@

\subsection*{3. Building phylogenies}
From the resulting matrices we generated two types of trees: the ``best" tree inferred from the ``complete" matrix and the ``missing-data" trees inferred from the 125 matrices with various amounts of missing data. The ``true" tree was used to generate the ``complete" matrix and reflects the ``true" evolutionary history in our simulations. The ``best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods. In real world situations, the ``true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \cite{rozen2005}). Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the ``best" tree, rather than the ``true" tree, as the ``best" tree is generally what we have to work with.

\paragraph*{Maximum Likelihood}
The ``best" tree and the ``missing-data" trees were inferred using RAxML v. 8.0.20 \cite{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model \cite{tavare1986} (default GTRGAMMA in RAxML v. 8.0.20  \cite{Stamatakis21012014}). For the morphological data, we used the implemented Markov \textit{k} state model \cite{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20 \cite{Stamatakis21012014}).
We used RAxML because it is heavily used in the literature for Maximum Likelihood tree inference (e.g. \cite{rouresite-specific2011,Bogdanowicz2012,springermacroevolutionary2012,O'Leary08022013,kellymolecular2014}) and is one of the fastest methods available \cite{Stamatakis01102008}.

To measure the support for each branch in our simulated phylogenies we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the ``complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \cite{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete" and 50 x 125 ``missing-data" matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50). On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \cite{pattengale2010many}. Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes). We performed this procedure to increase the resolution of our resulting trees.

\paragraph*{Bayesian}
The ``best" tree and the ``missing-data" trees were inferred using MrBayes v. 3.2.1 \cite{Ronquist2012mrbayes}. We partitioned the data to treat the molecular part as a non-codon DNA partition and the morphological part as a multi-state morphological partition. The molecular evolutionary history was inferred using the HKY model with a transition/transversion ratio of two \cite{douadycomparison2003} and a gamma distribution for the rate variation with four distinct categories (HKY + $\Gamma_4$). For the morphological data, we used the Markov \textit{k} state model \cite{lewisa2001}, with equal state frequency and a unique overall substitution rate ($\mu$) with four distinct rates categories (M\textit{k} + $\Gamma_4$). We chose these models to be consistent with the parameters used to generate the ``complete" matrix.

@@

Additionally, the number of morphological characters determines the size the matrix. This can affect our ability to recover the ``best" tree topology through: (1) the incongruence of phylogenetic signal among morphological and molecular data; and/or (2) homoplasy. The incongruence of phylogenetic signal between morphological and molecular data has previously been demonstrated to be more important in small morphological matrices \cite{wagner2000}. The size of our data matrices were constrained by the performance of our protocol: to reduce the computational time of our analysis to a reasonable level (150 CPU years), we ran our simulations on modestly-sized matrices of 1000 molecular characters and 100 morphological characters. Therefore, part of the decrease of the Normalised Robinson-Fould distance and the Normalised Triplets distance in our simulations could be due to conflicting phylogenetic signal among morphological and molecular data in our matrices (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). However, although these matrices are an order of magnitude smaller than some published matrices (e.g. \cite{springermacroevolutionary2012,nithe2013}, they are still within the size range of more modestly-sized empirical matrices (e.g. \cite{kellymolecular2014, sallam2011craniodental}). Therefore, our simulations reflect realistic parameters. Homoplasy, on the other hand, is expected to increase with an increase in the number of morphological characters \cite{wrightbayesian2014}. However, the use of probabilistic methods (i.e. Maximum Likelihood or Bayesian) and the M\textit{k} model \cite{lewisa2001} has been previously demonstrated to partially resolve this issue \cite{wrightbayesian2014}.

@@

\subsection*{Effects of tree inference methods}
Variation in our ability to recover the ``best" tree topology depends heavily on the tree inference method (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). For morphological data, previous studies have shown some superiority of probabilistic tree inference methods with simple evolutionary models such as the M\textit{k} model \cite{lewisa2001} over cladistic methods \cite{wrightbayesian2014} (but see \cite{spencerefficacy2013}). However, this is the first study, to our knowledge, to compare the performance of the M\textit{k} model \cite{lewisa2001} for recovering the ``best" tree topology using Maximum Likelihood and Bayesian methods in a Total Evidence framework. Our results show that the topology of the Bayesian consensus tree is always closer to the ``best" tree topology than the one of the Maximum Likelihood tree (Fig.~\ref{Fig_Results-global_perparam}). As described above, this is because the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree. This may also be because the ``best" Bayesian consensus trees are not completely resolved, thus will always be more similar to the ``missing data" trees than a completely resolved tree like the ``best" Maximum Likelihood tree. However, we minimized the probability of unresolved ``best" trees in our Bayesian analyses by only using datasets with strong phylogenetic signal (see Methods).

>>>>

\section*{Methods}
To explore how missing data in the morphological sections of Total Evidence matrices influences tree topology, we used the following protocol (note that we explain each step in detail below this general outline; Fig.~\ref{Fig_Outline}).
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true" tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete" matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). We call the resulting 125 matrices ``missing-data" matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete" matrix and from the 125 ``missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the ``best" tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data" trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best" tree to the ``missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

@@

In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $M_C$ removes columns across both living and fossil taxa's data. Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa. We first applied the parameters $M_L$ and $M_F$ on the matrix and then applied the $M_C$ parameter. Therefore, when 10\% data was missing for both $M_L$ and $M_F$, 10\% data was missing in the morphological part of the matrix. However, when applying the $M_C$ parameter with the same rate (10\%) the resulting matrix potentially had more than 10\% data missing.

@@

\subsection*{3. Estimating phylogenies}
From the resulting matrices we generated two types of trees: the ``best" tree inferred from the ``complete" matrix and the ``missing-data" trees inferred from the 125 matrices with various amounts of missing data. The ``true" tree was used to generate the ``complete" matrix and reflects the ``true" evolutionary history in our simulations. The ``best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods. In real world situations, the ``true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \cite{rozen2005}). Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the ``best" tree, rather than the ``true" tree, as the ``best" tree is generally what we have to work with.

\paragraph*{Maximum Likelihood}
The ``best" tree and the ``missing-data" trees were inferred using RAxML v. 8.0.20 \cite{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model \cite{tavare1986} (default GTRGAMMA in RAxML v. 8.0.20  \cite{Stamatakis21012014}). For the morphological data, we used the implemented M\textit{kv} model \cite{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20 \cite{Stamatakis21012014}).
We used RAxML because it is heavily used in the literature for Maximum Likelihood tree inference (e.g. \cite{rouresite-specific2011,Bogdanowicz2012,springermacroevolutionary2012,O'Leary08022013,kellymolecular2014}) and is one of the fastest methods available \cite{Stamatakis01102008}.

To measure the support for each branch in our simulated phylogenies we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the ``complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \cite{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete" and 50 x 125 ``missing-data" matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50). This step was implemented to make sure that the differences we observed in topologies (see below) where due to the amount of missing data for each parameters ($M_L$, $M_F$ and $M_C$) and not simply to low branch support that is likely to lead to different topologies. On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \cite{pattengale2010many}. Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes). We performed this procedure to increase the resolution of our resulting trees.

\paragraph*{Bayesian}
The ``best" tree and the ``missing-data" trees were inferred using MrBayes v. 3.2.1 \cite{Ronquist2012mrbayes}. We partitioned the data to treat the molecular part as a non-codon DNA partition and the morphological part as a multi-state morphological partition. The molecular evolutionary history was inferred using the HKY model with a transition/transversion ratio of two \cite{douadycomparison2003} and a gamma distribution for the rate variation with four distinct categories (HKY + $\Gamma_4$). For the morphological data, we used the M\textit{kv} model \cite{lewisa2001}, with equal state frequency and a unique overall substitution rate ($\mu$) with four distinct rates categories (M\textit{kv} + $\Gamma_4$). We chose these models to be consistent with the parameters used to generate the ``complete" matrix.

@@

Additionally, the number of morphological characters determines the size the matrix. This can affect our ability to recover the ``best" tree topology through: (1) the incongruence of phylogenetic signal among morphological and molecular data; and/or (2) homoplasy. The incongruence of phylogenetic signal between morphological and molecular data has previously been demonstrated to be more important in small morphological matrices \cite{wagner2000}. The size of our data matrices were constrained by the performance of our protocol: to reduce the computational time of our analysis to a reasonable level (150 CPU years), we ran our simulations on modestly-sized matrices of 1000 molecular characters and 100 morphological characters. Therefore, part of the decrease of the Normalised Robinson-Fould distance and the Normalised Triplets distance in our simulations could be due to conflicting phylogenetic signal among morphological and molecular data in our matrices (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). However, although these matrices are an order of magnitude smaller than some published matrices (e.g. \cite{springermacroevolutionary2012,nithe2013}, they are still within the size range of more modestly-sized empirical matrices (e.g. \cite{kellymolecular2014, sallam2011craniodental}). Therefore, our simulations reflect realistic parameters. Homoplasy, on the other hand, is expected to increase with an increase in the number of morphological characters \cite{wrightbayesian2014}. However, the use of probabilistic methods (i.e. Maximum Likelihood or Bayesian) and the M\textit{kv} model \cite{lewisa2001} has been previously demonstrated to partially resolve this issue \cite{wrightbayesian2014}.

@@

\subsection*{Effects of tree inference methods}
Variation in our ability to recover the ``best" tree topology depends heavily on the tree inference method (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). For morphological data, previous studies have shown some superiority of probabilistic tree inference methods with simple evolutionary models such as the M\textit{kv} model \cite{lewisa2001} over cladistic methods \cite{wrightbayesian2014} (but see \cite{spencerefficacy2013}). However, this is the first study, to our knowledge, to compare the performance of the M\textit{kv} model \cite{lewisa2001} for recovering the ``best" tree topology using Maximum Likelihood and Bayesian methods in a Total Evidence framework. Our results show that the topology of the Bayesian consensus tree is always closer to the ``best" tree topology than the one of the Maximum Likelihood tree (Fig.~\ref{Fig_Results-global_perparam}). As described above, this is because the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree. This may also be because the ``best" Bayesian consensus trees are not completely resolved, thus will always be more similar to the ``missing data" trees than a completely resolved tree like the ``best" Maximum Likelihood tree. However, we minimized the probability of unresolved ``best" trees in our Bayesian analyses by only using datasets with strong phylogenetic signal (see Methods).
